Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-26 _tex/manuscript.tex
Built site for gh-pages

For later use, we define at this point the matrices
\(\boldsymbol{\Lambda}\):

@@

The single-patient likelihood \(\ell\) in
equation~\ref{eq-single-patient-llh} depends on the spread parameters
shown in figure~\ref{fig-small-graph} via the transition matrix
\(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
time-priors. In this work, we will only differentiate between ``early''
(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, our
parameter space is:

@@

Together with the explicit modelling of the tumor's midline extension
\(\epsilon\) from section~\ref{sec-midline}, we now have a model that
may be capable of capturing the higher prevalence of contralateral
involvment that comes with tumors extending over the mid-sagittal line.

@@

Technically, the introduction of this additional random variable doubles
the space of the hidden states and therefore quadruples the size of the
transition matrix \(\mathbf{A}\). But since only the transition matrix
in case of midline extension depends on the lateralization of the tumor,
we implement it differently and thereby make the computation slightly
more efficient.

First, the overall probabilities to find a patient with a clearly
lateralized tumor or one that extends over the mid-sagittal line after
\(t\) time-steps are respectively given by

@@

Using this, it is straightforward to write down the joint probability of
ipsi- and contralalteral states in case of a tumor that does \emph{not}
extend over the mid-sagittal line similar to the marginalization in
equation~\ref{eq-bilateral-marginal}:

\begin{equation}\phantomsection\label{eq-joint-noext}{
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{False} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot (1 - p_\epsilon)^t \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t, \epsilon=\texttt{False} \right)
}\end{equation}

where for the contralateral evolution we used the transition matrix that
depends on the base spread parameters
\(b_v^{\text{c},\epsilon=\texttt{False}}\).

The case when midline extension is eventually present is more
complicated: We already marginalize over the exact time-step when the
tumor grows over the mid-sagittal line. But whenever that happens, we
also need to change the contralateral transition matrix to use the
increased spread rates from the tumor to the contralateral LNLs, given
by the linear mix in equation~\ref{eq-mixing}. We can achieve the
correct marginalization by iteratively building the joint
\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\)
from \(t=0\), where we know that all LNLs are healthy and the tumor is
lateralized:

\[
P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
\]

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The tumor was still lateralized at \(t=\tau\) and just grew over the
  midline. Under this scenario, the probability for having a midline
  extension at \(t=\tau+1\) is \(p_\epsilon\) and we use this to weight
  the contralateral state distribution that was so far evolved without
  increased contralateral spread.
\item
  The mid-sagittal line was already crossed by the tumor. In that case,
  the probability is 1 to remain in that lateralization state. Thus, for
  this case we simply add the same distribution we are computing, but
  from one time-step earlier, arriving at a recursion relation:
\end{enumerate}

\[
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}^{\text{c},\epsilon=\texttt{True}}
\]

Using this, we can write the joint of ipsi- and contralateral
involvement also for the case of mid-sagittal extension:

\begin{equation}\phantomsection\label{eq-joint-ext}{
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)
}\end{equation}

Subsequently, equation~\ref{eq-joint-noext} and
equation~\ref{eq-joint-ext} can be used to compute the likelihoods of
patients with the respective tumor lateralization. If, for some reason,
it is not known whether the tumor crossed the midline or not, we can
simply add the two distributions to marginalize over midline extension
as well.

In figure~\ref{fig-model-state-dist} we plot the state distribution for
a full midline model as two separate heatmaps. To keep it readable, this
example model only considers the LNLs II, III, and IV ipsi- and
contralaterally.

@@

\caption{\label{fig-model-state-dist}3D state distribution of a midline
model with 3 LNLs (II, III, and IV) in both sides of the neck.}

>>>>

For later use, we define at this point the matrices of the dsitributions
over all hidden states, given all time-steps \(\boldsymbol{\Lambda}\):

@@

The single-patient likelihood \(\ell\) in
equation~\ref{eq-single-patient-llh} depends on the spread parameters
shown in figure~\ref{fig-small-graph} via the transition matrix
\(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
time-priors. In this work, we will only differentiate between ``early''
(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, the
parameter space of the unilateral model is:

@@

Our parameter space has now expanded to

\begin{equation}\phantomsection\label{eq-bi-param-space}{
\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
}\end{equation}

which is less than double in size, compared to the unilateral model.
From these parameters, we now build three different transition matrices:
The unchanged \(\mathbf{A}_\text{i}\) for the ipsilateral side, one
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the contralateral
side that covers the progression as long as the tumor is lateralized,
and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in case of a tumor
that has crossed the mid-sagittal line.

@@

The overall probabilities to find a patient with a clearly lateralized
tumor or one that extends over the mid-sagittal line after \(t\)
time-steps are then respectively given by

@@

Using this, it is straightforward to write down the matrix of state
distributions for all time-steps, as in equation~\ref{eq-lambda-matrix}
covering the contralateral hidden state evolution:

\[
\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{False}} =
\left(
\begin{array}{r}
\boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{0\phantom{t_\text{max}}} \\
(1-p_\epsilon) \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{1\phantom{t_\text{max}}} \\
\hfill \vdots \hfill \\
(1-p_\epsilon)^{t_\text{max}} \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{t_\text{max}\phantom{0}} \\
\end{array}
\right)
\]

where we used the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) that depends on the
base spread parameters \(b_v^{\text{c},\epsilon=\texttt{False}}\).

The case when midline extension is eventually present is more
complicated: We already marginalize over the exact time-step when the
tumor grows over the mid-sagittal line. But whenever that happens, we
also need to change the contralateral transition matrix to use the
increased spread rates \(b_v^{\text{c}, \epsilon=\texttt{True}}\) from
the tumor to the contralateral LNLs, given by the linear mix in
equation~\ref{eq-mixing}. We can achieve the correct marginalization by
iteratively building the joint
\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).
We start from \(t=0\), where we know that all LNLs are healthy - i.e.,
the contralateral neck is in the state
\(\mathbf{X}_\text{c}=\boldsymbol{\xi}_0\) - and the tumor is
lateralized (\(\epsilon=\texttt{False}\)):

\[
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid t=0 \right) = \boldsymbol{\xi}_0
\]

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The tumor was still lateralized at \(t=\tau\) and just grew over the
  midline. Under this scenario, the probability for having a midline
  extension at \(t=\tau+1\) is given by \(p_\epsilon\). We use this to
  weight the contralateral state distribution that was so far evolved
  without increased contralateral spread.
\item
  The mid-sagittal line was already crossed by the tumor. In that case,
  the probability is 1 to remain in that lateralization state. Thus, for
  this case we simply add the same distribution we are computing, but
  from one time-step earlier, arriving at a recursion relation:
\end{enumerate}

\[
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
\]

We can collect the iteratively computed distributions for the midline
extension case to define the matrix over the states given all
time-steps, in analogy to equation~\ref{eq-lambda-matrix}:

\[
\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{True}} = \begin{pmatrix}
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 0 \right) \\
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 1 \right) \\
\vdots \\
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t_\text{max} \right) \\
\end{pmatrix}
\]

Using this, we can again write the joint of ipsi- and contralateral
involvement - now also for the case of mid-sagittal extension -
algebraically as before in equation~\ref{eq-bilateral-marginal-algebra}:

\[
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
\]

We can use this term to compute the likelihood of all patients with and
without midline extension separately. And if for some patients the
information of tumor lateralization is not available, we can simply sum
the above term once for \(\epsilon=\texttt{False}\) and once for
\(\epsilon=\texttt{True}\) to marginalize over the unknown variable.

The final parameter space of our extended model has now reached this
size:

\begin{equation}\phantomsection\label{eq-ext-param-space}{
\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.}, p_\epsilon \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
}\end{equation}

In figure~\ref{fig-model-state-dist} we plot the state distribution for
a full midline model as two separate heatmaps. To keep it manageable to
interpret, this example model only considers the LNLs II, III, and IV
ipsi- and contralaterally, resulting in
\(2 \times 2^3 \times 2^3 = 128\) distinct states.

@@

\caption{\label{fig-model-state-dist}3D state distribution of a midline
model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the
higher probability for contralateral involvement in the right subplot,
where the tumor extends over the mid-sagittal line.}
